### 2019-12-07 method/sec_4-db_struct.tex
Added Data Preprocess

- Module ID
  String field which describe the source of the data generated. Example of hec-hms, flow2d, weather-station etc
- Value Type
  - Scalar, Vector, Grid are the Value Types which are interested in the WDIAS.
- Location
  Location of the timeseries. All locations have unique String identifier called locationId. Further, there are two type of location types that is interested in the WDIAS.
  - Point locations - Contains a name which is human readable. lat and lon of the location on earth surface.
  \begin{lstlisting}[language=Python]
    {
        "locationId": "wdias-hanwella",
        "name": "Hanwella",
        "lat": 6.909722222,
        "lon": 80.08166667
    }
  \end{lstlisting}
  - Regualr Grid locations - Contains a name which is human readable. The grid presents by deviding into equal size cells. Thus, it need the number of rows and columns. 
  And the location of the firt cell and the width and height of it.
  \begin{lstlisting}[language=Python]
      {
        "locationId": "wdias_kelani_basin",
        "description": "Kelani Basin",
        "rows": 120,
        "columns": 139,
        "geoDatum": "Kandawala",
        "gridFirstCell": {
            "firstCellCenter": {
                "x": 397074.0,
                "y": 504875.0
            },
            "xCellSize": 250.0,
            "yCellSize": 250.0
        }
    }
\end{lstlisting}
  - Irregualr Grid locations (endpoints are available with in the WDIAS system, but skipped since it is out of the interest of the scope.)

@@

- MySQL
Relational Database Management System (RDBMS).
SQL database has fewer version of open source and it is also possible to gain better performance with commercial versions. 
Since WDIAS scope is to using open source tools, it is using MySQL as persistent database for storing timeseries metadata described above.
One instance of MySQL is using only by adapter-metadata service.
Other than that, another instance of MySQL is using for adapter-extension which will be describe in more detail \ref{data_preprocess}.
MySQL is used only for these services, those will only keep the consistency of the schema for timeseries metadata and extension metadata.
\ref{db_struct_redis} instance is using for caching the data from MySQL, since the MySQL database instances, itself not going to get much load.
Other than MySQL, there are many open source databases can be used to replace the same functionality.

- InfluxDB \cite{influxdbInfluxDBDocumentation}
Open timeseries DB with MIT license
Support SQL-Like query language
Clustering available with a commercial version
Among the other timeseries open source databases available, InfluxDB has higher number of usage and support. Database is written in Golang programming language,
and has higher performance. Other than InfluxDB, another Database is Elastic Search which is mainly using for indexing for searching.
With changing the configurations of the InfluxDB, it is possible to adapt as required. As mention in \cite{influxdbInfluxDBDocumentation}, 
InfluxDB can setup on a computer which will give the required performance. Going further, users can use commercial version of InfluxDB in order to gain more performance.

@@

- NetCDF \cite{unidataUnidataNetCDF}
Network Common Data Form
Self-describing, machine-independent data formats that support creation, access, and sharing of array-oriented scientific data
Support parallel file access
NetCDF is widely using in sciencetific domain due to it's capability of storing data in multi dementions easily and flexibility with storing many of the sciencetific data.
Python wrapper for NetCDF-C library using in the WDIAS system, and use by adapter-grid service.

- MongoDB
  MongoDB is a general purpose, document-based, distributed database
  Supports query operations on geo-spatial data \cite{mongodbMongoDBManual}
  Clustering and Sharding available for scalability and reliability
MongoDB geo-spatial features are using for support Geo queries in the WDIAS. Other than that, external timeseries metadata queries are served by the indexed timeseries metadata data.
If metadata not present in the adapter-query, if fetch data from the adapter-metadata, then indexed and cache in the MongoDB. Vise a versa, if new timeseries create in the adapter-metadata,
it trigger an event to adapter-query to index the new timeseries.
\ref{micro_y_axis_scaling} concept is using here by splitting into the multiple services based on the functionality and scale the system.

- Redis \cite{redisRedisDocumentation}
  Redis is an open source (BSD licensed), in-memory data structure store, used as a database, cache.
  Clustering is available for scalability
Redis uses as in-memory caching for fast access of repreative access data in adapter-metadata and adapter-extension.
Other than that, adapter-status uses Redis as key value pair store for getting the status for async handling requests such as storing Grid data.
In adapter-extension uses Redis's Publisher Subscriber capabilities in order to create a cron job in extension-scheduler when new extension create on \ref{data_preprocess}

>>>>

\subsection{Timeseries Key Attributes}
- Module ID
  String field which describe the source of the data generated. Example of hec-hms, flow2d, weather-station etc
- Value Type
  - Scalar, Vector, Grid are the Value Types which are interested in the WDIAS.
- Location
  Location of the timeseries. All locations have unique String identifier called locationId. Further, there are two type of location types that is interested in the WDIAS.
  - Point locations - Contains a name which is human readable. lat and lon of the location on earth surface.
  \begin{lstlisting}[language=Python]
    {
        "locationId": "wdias-hanwella",
        "name": "Hanwella",
        "lat": 6.909722222,
        "lon": 80.08166667
    }
  \end{lstlisting}
  - Regualr Grid locations - Contains a name which is human readable. The grid presents by deviding into equal size cells. Thus, it need the number of rows and columns. 
  And the location of the firt cell and the width and height of it.
  \begin{lstlisting}[language=Python]
      {
        "locationId": "wdias_kelani_basin",
        "description": "Kelani Basin",
        "rows": 120,
        "columns": 139,
        "geoDatum": "Kandawala",
        "gridFirstCell": {
            "firstCellCenter": {
                "x": 397074.0,
                "y": 504875.0
            },
            "xCellSize": 250.0,
            "yCellSize": 250.0
        }
    }
\end{lstlisting}
  - Irregualr Grid locations (endpoints are available with in the WDIAS system, but skipped since it is out of the interest of the scope.)

@@

\subsection{MySQL}
Relational Database Management System (RDBMS).
SQL database has fewer version of open source and it is also possible to gain better performance with commercial versions. 
Since WDIAS scope is to using open source tools, it is using MySQL as persistent database for storing timeseries metadata described above.
One instance of MySQL is using only by adapter-metadata service.
Other than that, another instance of MySQL is using for adapter-extension which will be describe in more detail \ref{data_preprocess}.
MySQL is used only for these services, those will only keep the consistency of the schema for timeseries metadata and extension metadata.
\ref{db_struct_redis} instance is using for caching the data from MySQL, since the MySQL database instances, itself not going to get much load.
Other than MySQL, there are many open source databases can be used to replace the same functionality.

\subsection{InfluxDB} \cite{influxdbInfluxDBDocumentation}
Open timeseries DB with MIT license
Support SQL-Like query language
Clustering available with a commercial version
Among the other timeseries open source databases available, InfluxDB has higher number of usage and support. Database is written in Golang programming language,
and has higher performance. Other than InfluxDB, another Database is Elastic Search which is mainly using for indexing for searching.
With changing the configurations of the InfluxDB, it is possible to adapt as required. As mention in \cite{influxdbInfluxDBDocumentation}, 
InfluxDB can setup on a computer which will give the required performance. Going further, users can use commercial version of InfluxDB in order to gain more performance.

@@

\subsection{netCDF} \cite{unidataUnidataNetCDF}
\label{sub:netcdf}
Network Common Data Form
Self-describing, machine-independent data formats that support creation, access, and sharing of array-oriented scientific data
Support parallel file access
NetCDF is widely using in sciencetific domain due to it's capability of storing data in multi dementions easily and flexibility with storing many of the sciencetific data.
Python wrapper for NetCDF-C library using in the WDIAS system, and use by adapter-grid service.

\subsection{MongoDB}
\label{sub:mongodb}
  MongoDB is a general purpose, document-based, distributed database
  Supports query operations on geo-spatial data \cite{mongodbMongoDBManual}
  Clustering and Sharding available for scalability and reliability
MongoDB geo-spatial features are using for support Geo queries in the WDIAS. Other than that, external timeseries metadata queries are served by the indexed timeseries metadata data.
If metadata not present in the adapter-query, if fetch data from the adapter-metadata, then indexed and cache in the MongoDB. Vise a versa, if new timeseries create in the adapter-metadata,
it trigger an event to adapter-query to index the new timeseries.
\ref{micro_y_axis_scaling} concept is using here by splitting into the multiple services based on the functionality and scale the system.

\subsection{Redis}
\label{sub:redis}
  Redis\cite{redisRedisDocumentation} is an open source (BSD licensed), in-memory data structure store, used as a database, cache.
  Clustering is available for scalability
Redis uses as in-memory caching for fast access of repreative access data in adapter-metadata and adapter-extension.
Other than that, adapter-status uses Redis as key value pair store for getting the status for async handling requests such as storing Grid data.
In adapter-extension uses Redis's Publisher Subscriber capabilities in order to create a cron job in extension-scheduler when new extension create on \ref{se:data_preprocess}

### 2019-12-07 method/sec_5-data_preprocess.tex
Added Data Preprocess

\subsubsection{Extension Handler}

\subsubsection{Extension Scheduler}

>>>>

\subsubsection{Extension Handler}
Extension Handler is responsible for triggering the correct Extension with metadata when there is any change on subscribed extension trigger's OnChange timeseries.
\begin{figure}[htp]
    \centering
    \includegraphics[width=0.6\textwidth]{method/data_preprocess/validation.jpg}
    \caption{Extension Handler}
    \label{fig:extension_handler}
\end{figure}

In further elaboration, when new data imported into one of adapter-scalar or adapter-vector, those notify the extension-handler. Then extension handler check whether there is any extension subscribed for particular timeseries, and trigger it via triggering an event on particular extensions.

In order to improve the performance of matching triggers against a timeseries, the WDIAS added two mechanisms. For the separation of service responsibilities, those functionalities reside in the adapter-extension and expose via an endpoint.
First, other than storing extension metadata in MySQL instance of adapter-extension, it also store the data in reverse lookup table called triggers. Using this reverse lookup table, for a given trigger\_type and timeseries, it can retrieve the extensions which are part of given timeseries. Thus the response which is received by the 
extension-handler only contains the extensions to be triggered. No additional filtering need to be done at the extension-handler service.
Second, cache the resulting trigger\_type and timeseries data in \acrshort{redis} database for fast access and avoid too much queries on \acrshort{mysql} database.

\subsubsection{Extension Scheduler}
Extension Handler is responsible for triggering the correct Extension with metadata at given time which is provided at the creation on extension trigger's with the value of OnTime cronjob string value.

In order to improve the performance of triggering cronjobs at given time schedules, the WDIAS added few optimization mechanisms.
Same as for extension-handler, the adapter-extension expose an endpoint to retrieve extension metadata grouped by cronjob time. Then use \acrshort{go} to implement the extension-handler to fast processing the extensions. After retrieve extension which are grouped by cronjob time; within the extension-handler, it schedules programatic cronjobs to trigger at given times. When each cronjob triggers at given time, it also contains the extensions to be triggered. Fetching extension data and creating cronjobs occur with long period of time in order to avoid heavy overhead.
When a new extension trigger created, those data push into a \acrshort{redis} List. Then pulling from the extension-handler in regular time intervals and create new cronjobs for mentioned cronjob time.
As mentioned above, in order to avoid the expansion of the programmatic cronjob list which are created for new extension triggers, all cronjobs fetch again in long period of time and flush the previously scheduled cronjobs.
The extension-handler should be run as a single entity for the atrocity, in order to avoid triggering same extension twice.

### 2019-12-07 method/sec_6-query.tex
Added Data Preprocess

- Explain adapter query
How MongoDB is using for Geo Spatial search.

<Search API>

>>>>

As described in \ref{sub:mongodb}, adapter-query is primarily using for support Geo-spatial queries. Other than that, as the primary service for searching for timeseries, it also supports for retrieval of timeseries metadata for external queries as well.

\subsubsection{Timeseries Metadata}
\begin{lstlisting}
GET <HOST_NAME>/metadata/timeseries/<TIMESERIES_ID>
\end{lstlisting}

\begin{lstlisting}
GET <HOST_NAME>/metadata/timeseries?<QUERY_STRING>
QUERY_STRING can be consist of zero to many 
moduleId(Optional) HEC-HMS
valueType(Optional) Scalar
parameterId(Optional) O.Precipitation
locationId(Optional) wdias_hanwella
timeseriesType(Optional) ExternalHistorical
timeStepId(Optional) each_15_min
\end{lstlisting}

\subsubsection{Timeseries Geo Queries}
